\chapter{Introduction}
\label{chap:introduction}
% ============================================================

% ============================================================
\section{Motivation}
\label{sec:motivation}
% ============================================================

Financial time series data is never fully observed.
Limit-order book feeds drop quotes during bursts of order cancellations.
Equity markets suspend trading when prices move too violently.
Over-the-counter bond markets leave entire yield-curve tenors unquoted
for days.
Implied volatility surfaces contain large regions where no liquid
option trade has occurred.
Macro and alternative data arrive at mixed, irregular frequencies, with
publication lags that create structural gaps at the daily level.
In every one of these settings a practitioner or a researcher faces
the same decision: what value should be attributed to the missing
observation, and how much uncertainty should be attached to that
attribution?

This question is central for market makers and risk managers.
For instance, an emerging markets trader could be asked to make a market for
an highy illiquid instrument that has not traded for several hours or days.
Depending on the instrument, several approaches are possible: for a bond,
she could interpolate the missing yield from the rest of the curve; for a
derivative, she could use her own pricing model and hope that the imputs
are not too stale.\footnote{Note that here she is faced by the same problem, i.e.
factors (as $\alpha$, $\beta$, $\rho$ in SABR) that may have not been observed in a some time.}
The simplest solution is to carry forward the last observed value,
but this can lead to systematic mispricing, underestimation of risk, and distortion
of correlations among assets.

We aim to take a different path, one that looks for solutions that minimize a
loss between the imputed value and the true value, and that preserve as much
as possible the statistical properties of the series.
In classical machine learning approach, we want to
focus on quality of inference rather than interpretability of the model.

Furthermore, and again motivated by practical use cases,
we insist on a \emph{causal} imputation: the imputed value
at time $t$ must be based only on information available up to time $t$,
and not on any future observation. A market maker does not have the luxury of
looking into a magical orb to see what the price will be in the next few seconds.
We formalise this constraint through the language of stochastic
filtrations, survey and taxonomise the algorithms, from classical
statistical methods to modern deep generative models,
and evaluate them empirically on
representative financial datasets.

% ============================================================
\section{The Causal Constraint}
\label{sec:intro_causal}
% ============================================================

Let us quickly formalise basic notation and the notion of causal imputation.

Following \cite{grud}, we denote a multivariate time series with $D$ variables of length $T$ as
$\mathbf{X}_T = (\mathbf{x}_1, \mathbf{x}_2, \ldots, \mathbf{x}_T)^\top \in \mathbb{R}^{T \times D}$,
where $\mathbf{x}_t \in \mathbb{R}^D$ represents the $t$-th observations
(a.k.a., measurements) of all variables and $x_t^d$ denotes the measurement
of $d$-th variable of $\mathbf{x}_t$. Let $s_t \in \mathbb{R}$ denote the
time-stamp when the $t$-th observation is obtained and we assume that the
first observation is made at time $t = 0$ $(s_1 = 0)$.

Let $(\Omega, \mathcal{G}, \Prob)$ be a probability space and let
$\mathbb{F} = \{\mathcal{F}_t\}_{t \ge 1}$ be a filtration
representing the information available to a market participant through
time.
An imputation $\hat{\mathbf{x}}_t^d$ of a missing value $\mathbf{x}_t^d$ is causal if and
only if it is $\mathcal{F}_t$-measurable: it may depend on the
history $\{\mathbf{X}_s, \mathbf{M}_s\}_{s \le t}$ of observed values and missingness
indicators up to time $t$, but on nothing later.

The theoretical benchmark for any causal imputation is the conditional
expectation
\begin{equation}
    \hat{\mathbf{x}}_t^* = \E\!\left[\mathbf{x}_t \mid \mathcal{F}_t\right],
    \label{eq:intro_optimal}
\end{equation}
which is the unique $\mathcal{F}_t$-measurable estimator minimising
mean squared error.
In the linear Gaussian state-space model this optimum is attained
exactly by the Kalman filter prediction step.
For the non-linear, non-Gaussian models that may better describe
financial data, it can be approximated by recurrent neural networks,
diffusion-based generative models, or transformer architectures.

Note the crucial importance of $\sigma (\hat{\mathbf{x}}_t)$ ddd

% ============================================================
\section{Use Cases and Datasets}
\label{sec:intro_usecases}
% ============================================================

We organise the empirical investigation around six use cases, each
representing a distinct type of financial time series and a distinct
pattern of missingness.

\paragraph{UC\,1 --- Tick and high-frequency data (LOBSTER / TAQ).}
In a limit-order book, quotes are removed when orders are cancelled
or modified.
The resulting gaps in the bid--ask spread and mid-price are
endogenous to price movement: cancellations cluster around volatile
episodes, making the mechanism missing not at random (MNAR).
The causal constraint is most binding here, since even a
sub-second lookahead constitutes inadmissible information for an
execution algorithm.
We use the LOBSTER dataset , which provides
full ten-level limit-order book snapshots for NASDAQ-listed equities
under an academic licence, and the Trade and Quote (TAQ) database
via WRDS for cross-validation.

\paragraph{UC\,2 --- Daily equity prices (CRSP / Yahoo Finance).}
Trading halts triggered by circuit breakers and regulatory
suspensions produce gaps of one to five days in the equity price
record.
Survivorship bias further distorts panels from which delisted stocks
have been removed.
The missingness is conditionally missing at random (MAR) during
ordinary halts but shifts to MNAR when the halt is caused by
undisclosed information.
We use the Center for Research in Security Prices (CRSP) daily
stock file via WRDS as the primary dataset, with Yahoo Finance
(\texttt{yfinance} API) as a freely accessible alternative.

\paragraph{UC\,3 --- Fixed income and yield curves (FRED / ECB SDW).}
Over-the-counter bond markets produce sparse and irregular quote
streams.
Off-the-run Treasuries and corporate bonds can go unquoted for
days, leaving holes in the yield curve that must be filled
causally for real-time risk systems.
The missingness pattern ranges from MCAR for the most liquid
on-the-run maturities to MNAR for illiquid off-the-run tenors,
where the absence of a quote is itself informative about the
prevailing liquidity regime.
We use the FRED daily yield-curve series (US Treasuries, freely
available via the FRED API) and the ECB Statistical Data Warehouse
(euro-area sovereign rates) as primary sources.

\paragraph{UC\,4 --- Macro and mixed-frequency data (FRED-MD).}
Macro series such as GDP, industrial production, and PMI are
published monthly or quarterly, creating structural gaps at the
daily frequency required by nowcasting and trading-signal models.
Sensor failures and API outages additionally introduce irregular
MCAR gaps.
The FRED-MD dataset , comprising 128
monthly US macroeconomic series, is widely used as a benchmark
in the nowcasting literature, enabling direct comparison with
published results.

\paragraph{UC\,5 --- FX and cryptocurrency (Binance / Dukascopy).}
Foreign-exchange and cryptocurrency markets operate around the clock,
yet liquidity is thin during certain hours or following market stress.
The resulting OHLCV gaps are MNAR: illiquidity is endogenous to the
bid--ask spread and realised volatility.
We use minute-level OHLCV data from the Binance public REST API
(cryptocurrency) and tick-level foreign-exchange data from the
Dukascopy free historical data service, both of which can be
downloaded programmatically without institutional access.

\paragraph{UC\,6 --- Implied volatility surfaces (CBOE / OptionMetrics).}
Far out-of-the-money and long-dated options are illiquid; many
(strike, maturity) nodes in the implied volatility surface go
unquoted on any given day.
Imputation here is a two-dimensional problem (strike and maturity)
with the additional constraint that any completed surface must
respect no-arbitrage conditions: put--call parity and the
convexity of option prices as a function of strike.
We use the CBOE VIX term-structure series as a freely available
proxy and the OptionMetrics IvyDB database via WRDS for the full
surface.

The six use cases span all three missingness mechanisms, gap lengths
from one tick to several weeks, and data frequencies from
millisecond to monthly.
For tractability, the main empirical chapters focus on
UC\,2 (daily equities), UC\,3 (yield curves), and UC\,5
(high-frequency FX/crypto), which collectively cover the most
practically important settings and the widest methodological range.
UC\,1 and UC\,6 are treated as extensions in the final chapter.

% ============================================================
\section{Objectives and Contributions}
\label{sec:contributions}
% ============================================================

The thesis makes the following contributions.

\begin{enumerate}

    \item \textbf{A unified causal framework.}
    We provide a formal definition of causal imputation under a
    stochastic filtration $\mathbb{F}$, prove the optimality of the
    conditional expectation $\E[X_t \mid \mathcal{F}_t]$, and
    introduce the \emph{cost of causality} $\Delta_t$ as the MSE
    gap between the causal optimum and the acausal smoother.
    This quantity is estimated empirically for each algorithm and
    dataset, providing a principled measure of how much predictive
    information resides in future observations.

    \item \textbf{A taxonomy of causal imputation algorithms.}
    We survey methods from na\"{i}ve baselines (LOCF) through
    classical statistical models (ARIMA, exponential smoothing),
    state-space filters (Kalman, EKF, particle filter), multiple
    imputation (EM, MICE), classical machine learning (kNN,
    missForest, Gaussian processes, low-rank completion), and
    deep learning (GRU-D ,
    BRITS 
    Neural ODEs 
    SAITS
    GP-VAE 
    CSDI 
    For each method we identify whether its standard form is causal,
    and where it is not, we derive or propose a causal variant.

    \item \textbf{A standardised benchmark under the causal constraint.}
    We define a rigorous evaluation protocol with a strict causality
    assumption in evaluation (no algorithm may access $Y_s$ with
    $s > t$ at test time), standardised missingness injection under
    MCAR, MAR, and MNAR at missing rates $\rho \in \{0.1, 0.3, 0.5\}$,
    and a common suite of metrics including RMSE, MAE, MRE, CRPS,
    and calibration coverage.

    \item \textbf{Empirical results on financial datasets.}
    We report skill scores relative to the LOCF baseline for all
    algorithms across UC\,2, UC\,3, and UC\,5, with ablations by
    gap length, missing rate, and missingness mechanism.
    Our results quantify the practical cost of the causal constraint
    and identify which algorithm classes offer the best trade-off
    between statistical performance and computational feasibility
    for real-time deployment.

\end{enumerate}

% ============================================================
\section{Thesis Structure}
\label{sec:structure}
% ============================================================

The remainder of the thesis is organised as follows.

\cref{chap:literature} reviews the missing-data and time-series
imputation literature, situating our work relative to foundational
contributions in statistics 

\cref{chap:framework} develops the mathematical framework:
probability-space foundations, Rubin's missingness taxonomy,
the formal definition of causal imputation, loss functions and
evaluation criteria, the state-space model and Kalman filter as the
theoretical gold standard, and extensions to non-linear and
non-Gaussian settings.

\cref{chap:algorithms} presents the full taxonomy of algorithms,
organised from na\"{i}ve baselines to deep generative models, with
a formal causal audit of each method's inference procedure.

\cref{chap:data} describes the datasets corresponding to UC\,2,
UC\,3, and UC\,5, the missingness-injection methodology, and the
train/validation/test splitting protocol that enforces the causal
constraint throughout evaluation.

\cref{chap:experiments} reports and discusses the empirical results,
including main comparisons, ablations, and an analysis of the
estimated cost of causality $\Delta_t$ across settings.

\cref{chap:conclusion} concludes with a summary of findings,
practical recommendations for practitioners, and directions for
future research, including the extension to UC\,1 (limit-order book)
and UC\,6 (implied volatility surface) and the open problem of
causal diffusion-based imputation under $\mathcal{F}_t$.

% ============================================================
% Bibliography entries specific to the introduction
% (all entries must be present in references.bib)
% ============================================================

% New entries to add to references.bib if not already present:
%   McCrackenNg2016, Huang2011LOBSTER, Fortuin2020GPVAE
% (see references.bib for existing entries cited above)